Temperature and moisture effects on greenhouse gas emissions from deep active-layer boreal soils

### 2016-11-20 manuscript/response_to_reviewer_comments.tex

\medskip
{\it 3.24. 393: 'taking them out of depth' rephrase. Also could use this argument for the section on CH4 production. }

We have reworded this. {\bf See new lines LINES.}

\medskip
{\it 3.25. Fig.1 : Edit figure to be color-blind friendly. }

We thought we were already doing so in using the {\tt RColorBrewer} package, not the default palette of {\tt ggplot2}, but have shifted to using a color-blind friendly palette from \url{http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/#a-colorblind-friendly-palette} in all figures.

### 2016-11-20 manuscript/cpcrw_manuscript.Rmd
Posted data, added DOI
Tweaks for concision and clarity

Rapid climatic changes, rising air temperatures, and increased fires are expected to drive permafrost degradation and alter soil carbon (C) cycling in many high-latitude ecosystems. How these soils will respond to changes in their temperature, moisture, and overlying vegetation is highly uncertain, but critical to understand given the large soil C stocks in these regions. We used a laboratory experiment to examine how temperature and moisture control CO~2~ and CH~4~ emissions from mineral soils sampled from the bottom of the annual active layer, i.e. directly above permafrost, in an Alaskan boreal forest. Gas emissions from thirty cores, subjected to two temperatures and either field moisture conditions or experimental drought, were tracked over a 100-day incubation; we also measured a variety of physical and chemical characteristics of the cores. Gravimetric water content was `r pn(gwcbegin$mean, 2)` ± `r pn(gwcbegin$sd, 2)` (unitless) at the beginning of the incubation; cores at field moisture were unchanged at the end, but drought cores had declined to `r pn(gwcendmean["Drought"], 2)` ±  `r pn(gwcendsd["Drought"], 2)`. Daily CO~2~ fluxes were positively correlated with incubation chamber temperature, core water content, and percent soil nitrogen, and had a temperature sensitivity (Q~10~) of `r pn(Q10_cumulative["Field moisture", "CO2"], 1)` and `r pn(Q10_cumulative["Drought", "CO2"], 1)` for the field moisture and drought treatments, respectively. Daily CH~4~ emissions were most strongly correlated with percent nitrogen, but neither temperature nor water content was a significant first-order predictor of CH~4~ fluxes. The cumulative production of C from CO~2~ was over `r minmag` orders of magnitudes higher than that from CH~4~; cumulative CO~2~ was correlated with incubation temperature and moisture treatment, with drought cores producing `r pn(drought_effect["20", "Reduction"], 0)`% - `r pn(drought_effect["4", "Reduction"], 0)`% lower C. Cumulative CH~4~ production was unaffected by any treatment. These results suggest that deep active-layer soils may be quite sensitive to changes in soil moisture under aerobic conditions, a critical factor as discontinuous permafrost thaws in interior Alaska. Deep but unfrozen high-latitude soils have been shown to be strongly affected by long-term experimental warming, and these results provide insight into their future dynamics and feedback potential with future climate change.

@@

High latitude ecosystems are being subjected to rapid changes in climate {IPCC, 2013 #3916} and increases in fire frequency and intensity {Kasischke, 2010 #2631}, notably in northwestern North America and Alaska {Hinzman, 2005 #4315}{Ju, 2016 #4320}. This will have a wide variety of ecosystem effects {Alexander, 2016 #4414}: in particular, rising temperatures and increasing fire will likely result in permafrost degradation {Pastick, 2015 #4311}{Zhang, 2015 #4372}{Genet, 2013 #4412}{Helbig, 2016 #4413} and changes in soil temperature, with subsequent hydrology changes that will influence soil greenhouse gas (GHG) fluxes to the atmosphere {Schädel, 2014 #4472}. Such fluxes are a large component of the global C cycle and, because of the high C stocks of northern soils {Tarnocai, 2009 #2231}, could result in a significant and positive climate feedback {Treat, 2015 #4217}{Koven, 2011 #3092}{Schaefer, 2014 #4117}.

The magnitude, timing, and form-in particular as methane (CH~4~) or carbon dioxide (CO~2~)-of such any such feedback remain highly uncertain {Schuur, 2015 #4319}. While northern soils hold enormous quantities {Hugelius, 2014 #3978} of potentially mineralizable soil organic carbon (SOC), vegetation and succession dynamics (for example, thermal insulation by mosses) promote permafrost resilience to even large temperature changes {Jorgenson, 2010 #2627}{Turetsky, 2012 #3405}. Vegetation type also influences SOC quality and quantity, with microbial communities {Högberg, 2007 #4482}, soil respiration {Raich, 2000 #596}, and SOC  all linked to aboveground factors such as woody versus nonwoody stems, deciduous versus evergreen canopies, and the presence of nitrogen-fixing plants.

A number of factors may however disrupt these these feedbacks between vegetation type, ground cover, permafrost, and SOC stability. Climate changes, in particular regional warming and drying, cause vegetation stress {Ju, 2016 #4320}{Barber, 2000 #1320} and increased mortality. Conversely, increasing plant productivity in some regions can stimulate the decomposition of older SOC {Hartley, 2012 #3481}. Climate also drives fire regime changes, and ecosystem disruption is particularly likely after more-intense fire events {Johnstone, 2010 #2465}{Genet, 2013 #4412}. Even absent disturbance, the stability of SOC is highly uncertain, as it depends on soil temperature and moisture, the ages of and ratio between the carbon (C) and nitrogen (N) pools {Weiss, 2015 #4317}{Karhu, 2014 #4124}, and its protection from competent microorganisms, enzymes, and resources {Bailey, 2012 #3624}, whether by organomineral sorption, chemical lability, or physical location {Schmidt, 2011 #3135}.

Temperature and moisture typically have strong and often interactive influences on soil GHG emissions. Laboratory incubations, field observations, and meta-analyses have documented changing greenhouse gas (GHG) fluxes with rising temperature {Olefeldt, 2013 #3456}{Davidson, 2006 #1669}{Hashimoto, 2015 #4231}{Treat, 2015 #4217}. GHG responses to wetting and thawing dynamics exhibit substantial variability between studies, probably due to differences in soil type, antecedent conditions, phase changes, experimental protocols, etc. {Kim, 2012 #2863}. The anaerobic conditions common following permafrost thaw are expected to lower CO~2~ emissions but increase those of CH~4~ {Treat, 2015 #4217;Treat, 2014 #4213}, but emissions from aerobic soils will likely dominate the permafrost C feedback {Schädel, 2016 #4444}. Decadal warming and drying trends in Alaska {Bieniek, 2014 #4389} thus seem likely to increase GHG emissions from soils, and incubation experiments are critical to understand these dynamics {Elberling, 2013 #3720}.

@@

The goal of this study was to examine how temperature and moisture control GHG (CO~2~ and CH~4~) emissions from soils sampled from the bottom of the annual active layer–i.e., directly above permafrost–in an Alaskan boreal forest. We also aimed to characterize the chemical and structural properties of these soils following a 100-day incubation at different temperatures, subjecting some cores to drying treatments. We hypothesized that (i) CO~2~ would be the dominant pathway for C loss in these largely aerobic soils; (ii) soils maintained at field moisture and high temperature would lose more C-CO~2~ than cores incubated at 4˚C, due to increased aerobic and anaerobic microbial activity; and (iii) core CH~4~ fluxes would be sensitive only to temperature, as no anaerobic conditions were imposed on the cores.

@@

We sampled soils from a southeast slope (65.1620 °N, 147.4874 °W) at CPCRW, in a 60 m transition zone between lowland *Picea mariana* and upland *Betula neoalaskana*, with significant white spruce (*Picea glauca*) presence as well. Stand density in this transition zone was 4060 ± 2310 trees ha^-1^, with basal area of 27.9 ± 7.0 m^2^ ha^-1^. The forest was at least 90 years old {cf. \\Morishita, 2014 #4316} according to tree rings taken at the stem base of several of the largest white spruce. The soil is characterized as a poorly-drained silt loam, and on average had ~20 cm of organic material over the mineral soil.

@@

On 18 August 2015 cores were placed into one of two growth chambers (Conviron Control Systems BDW80, Winnipeg, Canada) maintained at 4 and 20 °C temperatures and 70% relative humidity and allowed to equilibrate for two weeks. Starting on 31 August 2015 we measured the cores' mass and GHG (CH~4~ and CO~2~) emissions four times in the first week, then twice per week for the first month, and then once per week for the rest of the 100-day incubation. Throughout the incubation, cores had a 200 µm mesh screen fit to the base and were mounted on porous ceramic plates (Soil Moisture Equipment Corp., Santa Barbara, CA, USA) so that, when the plates were placed in contact with water, water would move up into the cores via capillary action. The "drought" cores were mounted on dry plates, but not allowed to drop below 5% water content. After each flux measurement, cores received additional wetting from the top to maintain their water status at the desired level.

@@

The incubation experiment concluded on 9 December 2015, following the final CO~2~ and CH~4~ readings. Each soil core was maintained at the treatment-dependent temperature and moisture content (by mass) until removed for destructive sampling, December 14-18, 2015. Sub-samples were collected and composited throughout each soil core for dissolved organic carbon analysis (110 ± 24 g dry mass equivalent) and dry-mass calculations (~28 g each). The remaining core material was air-dried and separated into particles (>2 mm diameter) and soil (≤2 mm) using a U.S. Standard Test Sieve No. 10 (Fisherbrand,  Pittsburg, PA, USA). The dry mass and volume of soil were used in calculations of gravimetric and volumetric soil moisture content, respectively {Gardner, 1986 #4345}. Soil volume was calculated as the total core volume minus the volume of particles >2 mm diameter, with the latter determined by water displacement. Air-dried soil and sub-samples stored at -20 °C were sent to the Agricultural and Environmental Services Laboratory at the University of Georgia Extension in February 2016 for total C, N, and dissolved organic C. Samples were combusted in an oxygen atmosphere at 1350 °C, and measured for gaseous C and N using a Elementar Vario Max CNS. DOC was measured using a Shimadzu 5000 TOC Analyzer.

@@

For each measurement of each sample throughout the 100-day incubation (i.e., each gas, core, and date/time), we used the rise in gas concentrations to calculate a flux rate in ppm s^-1^ (CO~2~) or ppb s^-1^ (CH~4~), a linear rate of change (δc/δt) based on the concentration rise from a minimum (up to `r MAX_MINCONC_TIME` seconds after measurement began) to a maximum (at `r MAX_MINCONC_TIME`-`r MAX_MAXCONC_TIME` seconds). Each core’s respiration flux (*F*) was then calculated as $F=\frac{\delta c}{\delta t} \frac{V}{M} \frac{P_a}{RT}$ where *V* is the core-specific system volume, *M* the core dry mass as determined at the end of the incubation, *P~a~* atmospheric pressure (101 kPa; the incubation chambers were ~120 m a.s.l.), *R* the universal gas constant (8.3 x 10^-3^ m^3^ kPa mol^-1^ K^-1^) and *T* the chamber air temperature (K) at time of measurement. The final respiration rate was expressed on a dry soil mass basis (µg C g soil^-1^ day^-1^).

@@

All data analysis and statistics were performed using `r R.version.string` {R Development Core Team, 2016 #4314}. This experiment was run as an 'open experiment' {Bond-Lamberty, 2016 #4388} with all analysis code, data (from raw instrument data to final summaries), diagnostics, etc., available at https://github.com/bpbond/cpcrw_incubation. The summarized flux data backing the main results have been archived at [*DOI to be filled in*].

@@

Carbon dioxide fluxes during the incubation ranged from `r pn(fluxsummary["flux_µgC_gC_day_min", "CO2"], 1)` µg C g C^-1^ day^-1^ to a maximum of `r pn(fluxsummary["flux_µgC_gC_day_max", "CO2"], 1)` µg C g C^-1^ day^-1^, with a mean of `r pn(fluxsummary["flux_µgC_gC_day_mean", "CO2"], 1)` µg C g C^-1^ day^-1^ over the 100 days. CH~4~ rates ranged from `r pn(fluxsummary["flux_µgC_gC_day_min", "CH4"]*1000, 2)` ng C g C^-1^ day^-1^ to a maximum of `r pn(fluxsummary["flux_µgC_gC_day_max", "CH4"]*1000, 2)` ng C g C^-1^ day^-1^, with a mean of `r pn(fluxsummary["flux_µgC_gC_day_mean", "CH4"]*1000, 2)` ng C g C^-1^ day^-1^.

These means conceal considerable variability over the course of the incubation (**Figures 2 and 3**). In the linear mixed-effects model (AIC = `r pn(AIC(step_co2_lme), 1)`), instantaneous CO~2~ flux was positively correlated with incubation chamber temperature, core gravimetric water content, and percent soil N (all P < 0.05, and the latter two P < 0.001; **Table 2**). Temperature sensitivity decreased significantly (`r pclean(co2_temp_time)`) over the course of the incubation, while moisture sensitivity was unaffected by time. Percent C and percent N were highly correlated (r = `r pn(cor(cndata$C_percent, cndata$N_percent), 2)`) for these cores. Because percent N was a slightly stronger predictor, it was retained in the model while percent C was excluded; cf. Colman and Schimel {, 2014 #4172}. The interaction between water content and percent N was also highly significant (`r pclean(summary(step_co2_lme)$tTable["WC_gravimetric:N_percent", "p-value"])`), although cores with N > 0.2\% exhibited little relationship between water content and CO~2~ flux (data not shown). Instantaneous CH~4~ fluxes were positively correlated with percent N, while water content exhibited significant interactions with percent N and DOC as a predictor (**Table 3**). This model had little predictive power (AIC = `r pn(AIC(step_ch4_lme), 1)`), however, and neither temperature nor water content was a significant first-order predictor of CH~4~ fluxes.

@@

The cumulative flux numbers above result in CO~2~ temperature sensitivity (Q~10~) values of `r pn(Q10_cumulative["Field moisture", "CO2"], 1)` and `r pn(Q10_cumulative["Drought", "CO2"], 1)` for the field moisture and drought treatments, respectively; the corresponding Q~10~ values  based on cumulative CH~4~ were `r pn(Q10_cumulative["Field moisture", "CH4"], 1)` and `r pn(Q10_cumulative["Drought", "CH4"], 1)`. Computing Q~10~ values based on fluxes normalized by water filled pore space changed these values only slightly: to `r pn(wfpsQ10_cumulative["Field moisture", "CO2"], 1)` and `r pn(wfpsQ10_cumulative["Drought", "CO2"], 1)` for CO~2~, respectively, and `r pn(wfpsQ10_cumulative["Field moisture", "CH4"], 1)` and `r pn(wfpsQ10_cumulative["Drought", "CH4"], 1)` for CH~4~.

@@

Rises in boreal air temperatures, and unpredictable precipitation changes, are changing fire disturbance regimes, warming and dry many soils, increasing vegetation stress, degrade permafrost and deepen the active layer {Schuur, 2015 #4319}, all with uncertain consequences for soil dynamics and GHG fluxes. In this laboratory experiment we found that CO~2~, but not CH~4~, fluxes from these oxic active-layer mineral soils were sensitive to temperature and, in particular, moisture.

A number of studies have measured microbial respiration and GHG fluxes very close to our study site. Morishita et al. {, 2014 #4316} quantified GHG fluxes at CPCRW and nearby forests, and found CO~2~ production to be correlated with both temperature and moisture in upland Cryosols, consistent with our results. They found however that CH~4~ uptake was driven by temperature only. Waldrop et al. {, 2010 #4380} incubated active-layer and permafrost soils from *Picea mariana* sites near Fairbanks, AK, observing aerobic Q~10~ values of 9.0 (active layer) and 2.3 (permafrost) from -5 to 5 °C, and flux rates of 0.001-0.10 µmol CH~4~ day^-1^ g^-1^ (~`r pn(0.001 * 16 / 0.012 / 1000, 3)`-`r pn(0.1 * 16 / 0.012 / 1000, 3)` ng C g C^-1^ day^-1^), and ~1-5 µg C-CO~2~ hr^-1^ g^-1^ (~`r pn(1 * 24 / 0.012, 0)`-`r pn(5 * 24 / 0.012, 0)` µg C g C^-1^ day^-1^), considerably higher than the CO~2~ rates observed here. During the first 100 days of an incubation of Fairbanks-area 0-10 cm mineral soils, Neff and Hooper {, 2002 #4390} observed fluxes of ~`r pn(0.2 * 1000 * 12/44, 0)`-`r pn(1.5 * 1000 * 12/44, 0)` µg C-CO~2~ g C^-1^ day^-1^, in line with the results here, while Wickland and Neff {, 2008 #1820} reported that temperature and moisture exhibited interactive and similar-magnitude effects on decomposition in *P. mariana* soils. Finally, Sierra et al. {, 2016 #4504} incubated homogenized organic soils from CPCRW and observed interactions between high temperatures, moisture, and oxygen concentration.

A number of synthesis studies have produced robust results documenting the dynamics and C feedback potential of Arctic and boreal soils; comparing to these results is useful because although the response of soil biota to stresses such as drought tends to differ between soil types, organisms, and vegetation, it is often broadly similar across biomes and climatic conditions {Manzoni, 2012 #3545}. Using two meta-analyses of aerobic and anaerobic permafrost soil incubations, Schädel et al. {, 2016 #4444} showed that C release was highly sensitive to temperature, and that soils released far more (220-520%) C under aerobic conditions. Our incubation was fully aerobic, but its results are consistent with the conclusion that respiration in the form of CO~2~ is likely to dominate the high latitude C feedback, and that aerobic soils, and the conditions under which currently waterlogged soils may drain, deserve particular attention. In terms of absolute flux rates, Treat et al. {, 2015 #4217} reported mean CO~2~ rates of 47 (all mineral soils) and 101 (for 20-100 cm soils) µg C-CO~2~ g C^-1^ day^-1^ from a pan-Arctic synthesis of anaerobic soil incubations, somewhat lower than our aerobic incubation results. Treat et al. {, 2014 #4213} also found CO~2~ and CH~4~ emissions to be strongly correlated with temperature and moisture based on an incubation of Alaskan peats. Whether climate change makes northern regions wetter or driers is thus a critical factor affecting the quantity and form of C release.

>>>>

Rapid climatic changes, rising air temperatures, and increased fires are expected to drive permafrost degradation and alter soil carbon (C) cycling in many high-latitude ecosystems. How these soils will respond to changes in their temperature, moisture, and overlying vegetation is uncertain, but critical to understand given the large soil C stocks in these regions. We used a laboratory experiment to examine how temperature and moisture control CO~2~ and CH~4~ emissions from mineral soils sampled from the bottom of the annual active layer, i.e. directly above permafrost, in an Alaskan boreal forest. Gas emissions from thirty cores, subjected to two temperatures and either field moisture conditions or experimental drought, were tracked over a 100-day incubation; we also measured a variety of physical and chemical characteristics of the cores. Gravimetric water content was `r pn(gwcbegin$mean, 2)` ± `r pn(gwcbegin$sd, 2)` (unitless) at the beginning of the incubation; cores at field moisture were unchanged at the end, but drought cores had declined to `r pn(gwcendmean["Drought"], 2)` ±  `r pn(gwcendsd["Drought"], 2)`. Daily CO~2~ fluxes were positively correlated with incubation chamber temperature, core water content, and percent soil nitrogen, and had a temperature sensitivity (Q~10~) of `r pn(Q10_cumulative["Field moisture", "CO2"], 1)` and `r pn(Q10_cumulative["Drought", "CO2"], 1)` for the field moisture and drought treatments, respectively. Daily CH~4~ emissions were most strongly correlated with percent nitrogen, but neither temperature nor water content was a significant first-order predictor of CH~4~ fluxes. The cumulative production of C from CO~2~ was over `r minmag` orders of magnitudes higher than that from CH~4~; cumulative CO~2~ was correlated with incubation temperature and moisture treatment, with drought cores producing `r pn(drought_effect["20", "Reduction"], 0)`% - `r pn(drought_effect["4", "Reduction"], 0)`% lower C. Cumulative CH~4~ production was unaffected by any treatment. These results suggest that deep active-layer soils may be sensitive to changes in soil moisture under aerobic conditions, a critical factor as discontinuous permafrost thaws in interior Alaska. Deep but unfrozen high-latitude soils have been shown to be strongly affected by long-term experimental warming, and these results provide insight into their future dynamics and feedback potential with future climate change.

@@

High latitude ecosystems are being subjected to rapid changes in climate {IPCC, 2013 #3916} and increases in fire frequency and intensity {Kasischke, 2010 #2631}, notably in northwestern North America and Alaska {Hinzman, 2005 #4315}{Ju, 2016 #4320}. This will have a wide variety of ecosystem effects {Alexander, 2016 #4414}: in particular, rising temperatures and increasing fire will likely result in changes in soil temperature and permafrost degradation {Pastick, 2015 #4311}{Zhang, 2015 #4372}{Genet, 2013 #4412}{Helbig, 2016 #4413}, with subsequent hydrology changes that will influence soil greenhouse gas (GHG) fluxes to the atmosphere {Schädel, 2014 #4472}. Such fluxes are a large component of the global C cycle and could result in a significant and positive climate feedback {Treat, 2015 #4217}{Koven, 2011 #3092}{Schaefer, 2014 #4117}.

The magnitude, timing, and form-in particular as methane (CH~4~) or carbon dioxide (CO~2~)-of such any such feedback remain highly uncertain {Schuur, 2015 #4319}. While northern soils hold enormous quantities {Hugelius, 2014 #4162} of potentially mineralizable soil organic carbon (SOC), vegetation and succession dynamics (for example, thermal insulation by mosses) promote permafrost resilience to even large temperature changes {Jorgenson, 2010 #2627}{Turetsky, 2012 #3405}. Vegetation type also influences SOC quality and quantity, with microbial communities {Högberg, 2007 #4482}, soil respiration {Raich, 2000 #596}, and SOC  all linked to aboveground factors such as woody versus nonwoody stems, deciduous versus evergreen canopies, and the presence of nitrogen-fixing plants. A number of factors may however disrupt these these feedbacks between vegetation type, ground cover, permafrost, and SOC stability. Climate changes, in particular regional warming and drying, cause vegetation stress {Ju, 2016 #4320}{Barber, 2000 #1320} and increased mortality. Conversely, increasing plant productivity in some regions can stimulate the decomposition of older SOC {Hartley, 2012 #3481}. Climate also drives fire regime changes, and ecosystem disruption is particularly likely after intense fires {Johnstone, 2010 #2465}{Genet, 2013 #4412}. Even absent disturbance, the stability of SOC is highly uncertain, as it depends on soil temperature and moisture, the ages of and ratio between the carbon (C) and nitrogen (N) pools {Weiss, 2015 #4317}{Karhu, 2014 #4124}, and its protection (whether by organomineral sorption, chemical lability, or physical location) from competent microorganisms, enzymes, and resources {Bailey, 2012 #3624}{Schmidt, 2011 #3135}.

Temperature and moisture typically have strong and often interactive influences on soil GHG emissions. Laboratory incubations, field observations, and meta-analyses have documented changing greenhouse gas (GHG) fluxes with rising temperature {Olefeldt, 2013 #3456}{Davidson, 2006 #1669}{Hashimoto, 2015 #4231}{Treat, 2015 #4217}. GHG responses to wetting and thawing dynamics exhibit substantial variability between studies, probably due to differences in soil type, antecedent conditions, phase changes, experimental protocols, etc. {Kim, 2012 #2863}. The anaerobic conditions common following permafrost thaw are expected to lower CO~2~ emissions but increase those of CH~4~ {Treat, 2015 #4217;Treat, 2014 #4213}, but emissions from aerobic soils will likely dominate the permafrost C feedback {Schädel, 2016 #4444}. Decadal warming and drying trends in Alaska {Bieniek, 2014 #4389} thus seem likely to increase GHG emissions from soils, and laboratory incubation experiments are critical to understand these dynamics {Elberling, 2013 #3720}.

@@

The goal of this study was to examine how temperature and moisture control GHG (CO~2~ and CH~4~) emissions from soils sampled from the bottom of the annual active layer–i.e., directly above permafrost–in an Alaskan boreal forest. We also aimed to characterize the chemical and structural properties of these soils following a 100-day incubation at different temperatures, subjecting some cores to drying treatments. We hypothesized that (i) CO~2~ would be the dominant pathway for C loss in these largely aerobic soils; (ii) soils maintained at field moisture and high (20˚C) temperature would lose more C-CO~2~ than cores incubated at 4˚C, due to increased aerobic and anaerobic microbial activity; and (iii) core CH~4~ fluxes would be small and sensitive only to temperature, as no anaerobic conditions were imposed on the cores.

@@

We sampled soils from a southeast slope (65.1620 °N, 147.4874 °W) at CPCRW, in a 60 m transition zone between lowland *Picea mariana* and upland *Betula neoalaskana*, with significant white spruce (*Picea glauca*) presence as well. Stand density in this transition zone was 4060 ± 2310 trees ha^-1^, with basal area of 27.9 ± 7.0 m^2^ ha^-1^. The forest was at least 90 years old {cf. \\Morishita, 2014 #4316} according to tree cores taken at the base of several of the largest white spruce. The soil is characterized as a poorly-drained silt loam, and on average had ~20 cm of organic material over the mineral soil.

@@

On 18 August 2015 cores were placed into one of two growth chambers (Conviron Control Systems BDW80, Winnipeg, Canada) maintained at 4 and 20 °C temperatures and 70% relative humidity and allowed to equilibrate for two weeks. Starting on 31 August 2015 we measured the cores' mass and GHG emissions four times in the first week, then twice per week for the first month, and then once per week for the rest of the 100-day incubation. Throughout the incubation, cores had a 200 µm mesh screen fit to the base and were mounted on porous ceramic plates (Soil Moisture Equipment Corp., Santa Barbara, CA, USA) so that, when the plates were placed in contact with water, water would move up into the cores via capillary action. The "drought" cores were mounted on dry plates, but not allowed to drop below 5% water content. After each flux measurement, cores received additional wetting from the top to maintain their desired water status.

@@

The incubation experiment concluded on 9 December 2015, following the final CO~2~ and CH~4~ readings. Each soil core was maintained at the treatment-dependent temperature and moisture content (by mass) until removed for destructive sampling, December 14-18, 2015. Sub-samples were collected and composited throughout each soil core for dissolved organic carbon analysis (110 ± 24 g dry mass equivalent) and dry-mass calculations (~28 g each). The remaining core material was air-dried and separated into particles (>2 mm diameter) and soil (≤2 mm) using a U.S. Standard Test Sieve No. 10 (Fisherbrand,  Pittsburg, PA, USA). The dry mass and volume of soil were used in calculations of gravimetric and volumetric soil moisture content, respectively {Gardner, 1986 #4345}. Soil volume was calculated as the total core volume minus the volume of particles >2 mm diameter, with the latter determined by water displacement. Air-dried soil and sub-samples stored at -20 °C were sent to the Agricultural and Environmental Services Laboratory at the University of Georgia Extension in February 2016 for total C, N, and dissolved organic carbon (DOC). Samples were combusted in an oxygen atmosphere at 1350 °C, and measured for gaseous C and N using a Elementar Vario Max CNS. DOC was measured using a Shimadzu 5000 TOC Analyzer.

@@

For each measurement of each sample throughout the 100-day incubation (i.e., each gas, core, and date/time), we used the rise in gas concentrations to calculate a flux rate in ppm s^-1^ (CO~2~) or ppb s^-1^ (CH~4~), a linear rate of change (δc/δt) based on the concentration rise from a minimum (up to `r MAX_MINCONC_TIME` seconds after measurement began) to a maximum (at `r MAX_MINCONC_TIME`-`r MAX_MAXCONC_TIME` seconds). Each core’s respiration flux (*F*) was then calculated as $F=\frac{\delta c}{\delta t} \frac{V}{M} \frac{P_a}{RT}$ where *V* is the core-specific system volume, *M* the core dry mass as determined at the end of the incubation, *P~a~* atmospheric pressure (101 kPa; the incubation chambers were ~120 m a.s.l.), *R* the universal gas constant (8.3 x 10^-3^ m^3^ kPa mol^-1^ K^-1^) and *T* the chamber air temperature (K) at time of measurement. The final respiration rate was expressed on a soil C basis (µg or ng C g C^-1^ day^-1^).

@@

All data analysis and statistics were performed using `r R.version.string` {R Development Core Team, 2016 #4314}. This experiment was run as an 'open experiment' {Bond-Lamberty, 2016 #4388} with all analysis code, data (from raw instrument data to final summaries), diagnostics, etc., available at https://github.com/bpbond/cpcrw_incubation. The summarized flux data backing the main results have been archived under the Digital Object Identifier m9.figshare.4240436.v1.

@@

Carbon dioxide fluxes during the incubation ranged from `r pn(fluxsummary["flux_µgC_gC_day_min", "CO2"], 1)` µg C g C^-1^ day^-1^ to a maximum of `r pn(fluxsummary["flux_µgC_gC_day_max", "CO2"], 1)` µg C g C^-1^ day^-1^, with a mean of `r pn(fluxsummary["flux_µgC_gC_day_mean", "CO2"], 1)` µg C g C^-1^ day^-1^ over the 100 days. CH~4~ rates ranged from `r pn(fluxsummary["flux_µgC_gC_day_min", "CH4"]*1000, 2)` ng C g C^-1^ day^-1^ to a maximum of `r pn(fluxsummary["flux_µgC_gC_day_max", "CH4"]*1000, 2)` ng C g C^-1^ day^-1^, with a mean of `r pn(fluxsummary["flux_µgC_gC_day_mean", "CH4"]*1000, 2)` ng C g C^-1^ day^-1^. These means conceal considerable variability over the course of the incubation (**Figures 2 and 3**).

In the linear mixed-effects model (AIC = `r pn(AIC(step_co2_lme), 1)`), instantaneous CO~2~ flux was positively correlated with incubation chamber temperature, core gravimetric water content, and percent soil N (all P < 0.05, and the latter two P < 0.001; **Table 2**). Temperature sensitivity decreased significantly (`r pclean(co2_temp_time)`) over the course of the incubation, while moisture sensitivity was unaffected by time. Percent C and percent N were highly correlated (r = `r pn(cor(cndata$C_percent, cndata$N_percent), 2)`) for these cores. Because percent N was a slightly stronger predictor, it was retained in the model while percent C was excluded; cf. Colman and Schimel {, 2014 #4172}. The interaction between water content and percent N was also highly significant (`r pclean(summary(step_co2_lme)$tTable["WC_gravimetric:N_percent", "p-value"])`), although cores with N > 0.2\% exhibited little relationship between water content and CO~2~ flux (data not shown). Instantaneous CH~4~ fluxes were positively correlated with percent N, while water content exhibited significant interactions with percent N and DOC as a predictor (**Table 3**). This model had little predictive power (AIC = `r pn(AIC(step_ch4_lme), 1)`), however, and neither temperature nor water content was a significant first-order predictor of CH~4~ fluxes.

@@

The cumulative flux numbers above result in CO~2~ temperature sensitivity (Q~10~) values of `r pn(Q10_cumulative["Field moisture", "CO2"], 1)` and `r pn(Q10_cumulative["Drought", "CO2"], 1)` for the field moisture and drought treatments, respectively; the corresponding Q~10~ values  based on cumulative CH~4~ were `r pn(Q10_cumulative["Field moisture", "CH4"], 1)` and `r pn(Q10_cumulative["Drought", "CH4"], 1)`. Computing Q~10~ values based on fluxes normalized by water filled pore space changed these values only slightly: to `r pn(wfpsQ10_cumulative["Field moisture", "CO2"], 1)` and `r pn(wfpsQ10_cumulative["Drought", "CO2"], 1)` for CO~2~, for the field moisture and drought treatments respectively, and `r pn(wfpsQ10_cumulative["Field moisture", "CH4"], 1)` and `r pn(wfpsQ10_cumulative["Drought", "CH4"], 1)` for CH~4~.

@@

Rises in boreal air temperatures, and unpredictable precipitation changes, will change fire disturbance regimes, warm and dry many soils, increase vegetation stress, degrade permafrost, and deepen the active layer {Schuur, 2015 #4319}, all with uncertain consequences for soil dynamics and GHG fluxes. In this laboratory experiment we found that CO~2~, but not CH~4~, fluxes from these oxic active-layer mineral soils were sensitive to temperature and, in particular, moisture.

A number of studies have measured microbial respiration and GHG fluxes very close to our study site. Morishita et al. {, 2014 #4316} quantified GHG fluxes at CPCRW and nearby forests, and found CO~2~ production to be correlated with both temperature and moisture in upland Cryosols, consistent with our results. Waldrop et al. {, 2010 #4380} incubated active-layer and permafrost soils from *Picea mariana* sites near Fairbanks, AK, observing aerobic Q~10~ values of 9.0 (active layer) and 2.3 (permafrost) from -5 to 5 °C, and flux rates of 0.001-0.10 µmol CH~4~ day^-1^ g^-1^ (~`r pn(0.001 * 16 / 0.012 / 1000, 3)`-`r pn(0.1 * 16 / 0.012 / 1000, 3)` ng C g C^-1^ day^-1^), and ~1-5 µg C-CO~2~ hr^-1^ g^-1^ (~`r pn(1 * 24 / 0.012, 0)`-`r pn(5 * 24 / 0.012, 0)` µg C g C^-1^ day^-1^), considerably higher than the CO~2~ rates observed here. During the first 100 days of an incubation of Fairbanks-area 0-10 cm mineral soils, Neff and Hooper {, 2002 #4390} observed fluxes of ~`r pn(0.2 * 1000 * 12/44, 0)`-`r pn(1.5 * 1000 * 12/44, 0)` µg C-CO~2~ g C^-1^ day^-1^, in line with the results here, while Wickland and Neff {, 2008 #1820} reported that temperature and moisture exhibited interactive effects, of similar magnitude, on decomposition in *P. mariana* soils.

A number of synthesis studies have documented dynamics and C feedback potential of Arctic and boreal soils more generally; comparing to these results is useful because although the response of soil biota to stresses such as drought tends to differ between soil types, organisms, and vegetation, it is often broadly similar across biomes and climatic conditions {Manzoni, 2012 #3545}. Using two meta-analyses of aerobic and anaerobic permafrost soil incubations, Schädel et al. {, 2016 #4444} showed that C release was highly sensitive to temperature, and that soils released far more (220-520%) C under aerobic conditions. Our incubation was fully aerobic, but its results are consistent with the conclusion that respiration in the form of CO~2~ is likely to dominate the high latitude C feedback, and that aerobic soils, and the conditions under which currently waterlogged soils may drain, deserve particular attention. In terms of absolute flux rates, Treat et al. {, 2015 #4217} reported mean CO~2~ rates of 47 (all mineral soils) and 101 (for 20-100 cm soils) µg C-CO~2~ g C^-1^ day^-1^ from a pan-Arctic synthesis of anaerobic soil incubations, somewhat lower than our aerobic incubation results. Treat et al. {, 2014 #4213} also found CO~2~ and CH~4~ emissions to be strongly correlated with temperature and moisture based on an incubation of Alaskan peats. Whether climate change makes northern regions wetter or driers is thus a critical factor affecting the quantity and form of C release.

The drought treatment imposed in this experiment reduced soil C fluxes by `r pn(drought_effect["20", "Reduction"], 0)`% - `r pn(drought_effect["4", "Reduction"], 0)`%. The importance of this result depends, in part, on the spatial extent and intensity of precipitation changes across the boreal and Arctic this century. There is a detectable anthropogenic influence in high latitude precipitation changes {Wan, 2015 #4478}, but these changes are inconsistent: drier and warmer conditions in boreal Eurasia {Buermann, 2014 #3992}, for example, but growing season length increases in interior Alaska with no increase in precipitation {Wendler, 2009 #4477}. This spatial variability will interact with permafrost thaw dynamics to produce a complex patchwork of soil moisture changes {Zhang, 2012 #4479}{Watts, 2012 #4505}. The high uncertainty in this area makes it all the more important to understand the interactive effects of soil moisture and temperature on decomposition and GHG emissions {Sierra, 2015 #4203}.
